Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-31 paper/ijm.tex
Minor spelling mistakes corected

However, in case of less-than-perfect ethnic integration, the ``local''
language in ethnic enclaves may be the minority language instead, in
this way limiting the usefulness of the majority language skills
\citep{chiswick+miller2002,hwang+2010} in particular in context of ethnic segregation and
limited access to upper-end jobs \citep{Toomet2011}. However, the
existing evidence suggests that the majority language is
an important production factor in the ethnic enclaves as well
\citep{zhou+logan1989, clark+drinkwater2000}.

@@

We find that the Estonian language, the majority and the official
language of the country, is associated with a very
small albeit slowly increasing wage premium over
the years. Interestingly, this is true only for women, Russian men enjoy
virtually no premium from the majority language skills.
Russian skills, on the other hand, are clearly related to
higher wage but the relationship is slowly fading. English fluency was
associated with a very large wage premium during the early years of
economic transition, later its importance
declined but remained sizeable. 
In contrast, the sole language that plays a role in determining
unemployment probability is Estonian. Its fluency is associated with 
approximately 5 percentage points lower probability of being unemployed.
These results suggest that the Estonian labour market is 
segmented both along gender and the along ethnic background, with men
more likely occupied in less communication-intense jobs.

@@

We document that all analysed language skills are associated with
unemployment and wage, however in a different way. Our most
intriguing observation is that the largest group of
Russian men, high school graduates, earn virtually no income premium if fluent in Estonian,
the majority and the official language of the country, while ability
to speak the language is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015}, and \citet{Lindemann2013} findings that 
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent in the male
labour market.
This suggests for the majority of Russian speaking men, Estonian
language gives a distinct advantage by facilitating hiring by
Estonian-dominated employers. We also confirm
\citet{leppik+vihalemm2015JofBaltStud} conclusion that
these
jobs are associated with little career mobility. There is also some
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while manufacturing units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.
On a more positive tone, we find
that this may not be the case for highly educated workers.

The ethnic segregation seems to be less of an issue for women.
Unfortunately, neither \citet{Toomet2011} nor
\citet{leppik+vihalemm2015JofBaltStud} analyse men and women
separately. We find some evidence that female labour market is less
segregated along ethnic lines, and women are more frequently employed in jobs in which they need Estonian
language more, such as working in a direct contact with customers.
We also find evidence that over time Estonian is becoming more
important in the female labour market. Unfortunately, the current data
does not let us assess whether is due to decreasing segregation, or
economy--wide rise of the importance of Estonian language. No such trend is visible for
men suggesting that 20 years ---almost a generation--- after
establishing a nation state and market economy, the
labour market integration is still not complete.

>>>>

However, in the case of less-than-perfect ethnic integration, the ``local''
language in ethnic enclaves may be the minority language instead, in
this way limiting the usefulness of the majority language skills
\citep{chiswick+miller2002,hwang+2010} in particular in context of ethnic segregation and
limited access to upper-end jobs \citep{Toomet2011}. However, the
existing evidence suggests that the majority language is
an important production factor in the ethnic enclaves as well
\citep{zhou+logan1989, clark+drinkwater2000}.

@@

We find that the Estonian language, the ethnic majority and official
language of the country, is associated with a very
small albeit slowly increasing wage premium over
the years. Interestingly, this is true only for women, Russian men enjoy
virtually no premium from the majority language skills.
Russian skills, on the other hand, are clearly related to
higher wage but the relationship is slowly fading. English fluency was
associated with a very large wage premium during the early years of
economic transition, later its importance
declined but remained sizeable. 
In contrast, the sole language that plays a role in determining
unemployment probability is Estonian. Its fluency is associated with 
approximately 5 percentage points lower probability of being unemployed.
These results suggest that the Estonian labour market is 
segmented both along gender and the along ethnic background, with men
more likely occupied in less communication-intense jobs.

@@

We document that all analysed language skills are associated with
unemployment and wage, however in a different way. Our most
intriguing observation is that the largest group of
Russian men, high school graduates, earn virtually no income premium if fluent in Estonian,
the majority and the official language of the country, while ability
to speak the language is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015}, and \citet{Lindemann2013} findings that 
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent in the male
labour market.
This suggests for the majority of Russian-speaking men, Estonian
language gives a distinct advantage by facilitating hiring by
Estonian-dominated employers. We also confirm
\citet{leppik+vihalemm2015JofBaltStud} conclusion that
these
jobs are associated with little career mobility. There is also some
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while manufacturing units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.
On a more positive tone, we find
that this may not be the case for highly educated workers.

The ethnic segregation seems to be less of an issue for women.
Unfortunately, neither \citet{Toomet2011} nor
\citet{leppik+vihalemm2015JofBaltStud} analyse men and women
separately. We find some evidence that female labour market is less
segregated along ethnic lines, and women are more frequently employed in jobs in which they need Estonian
language more, such as working in a direct contact with customers.
We also find evidence that over time Estonian is becoming more
important in the female labour market. Unfortunately, the current data
does not let us assess whether is due to decreasing segregation, or
economy--wide rise of the importance of Estonian language. No such trend is visible for
men suggesting that 20 years ---almost a generation--- after
establishing a nation-state and market economy, the
labour market integration is still not complete.
